Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-05-02 ion_pairing_paper.tex
Simpler figure to describe procession

\begin{figure}
\includegraphics{dissociation.pdf}
\caption{Processionary model. (a) Sketch of the network in the case of a procession of size $n=3$ between cross-links (gray disks) and a persistence length $r_\mathrm{src}\approx 3D$. Empty circles are not individual monomers but electrostatic blobs of size $D$. Electrostatic blobs containing an anionic head are shown as black filled circles. (b) Relation between the number of monomers between uncondensed charges and the number of chains between cross-link points. The solid line corresponds to $n = A^2$. Full circles are obtained from (Eq.~\ref{eq:A}). In the case of \ce{PIm+Br-}, we are not in the validity regime of (Eq.~\ref{eq:A}), and we have to extrapolate the solid line to estimate $A$ (dotted arrow and open symbol).}
\label{fig:dissociation}
\end{figure}

@@

\begin{table}
\sisetup{
separate-uncertainty,
%table-format = 1.3(1),
}
\begin{tabular}{l|SSS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_0$} & {$\gamma_c$} & {$n$} & {$B$} & {$A$} & {$E_c$} & {$E_c$}\\
&	{\si{\pascal}} & {\%} & {\%} & & & & {\si{\kilo\joule\per\mole}}& {$k_\mathrm{B}T$}\\\hline&&&&&\\[-10pt]
\ce{PIm+Br-}	& 13	&	1.1	&	810	&	860	&	930	&	383	&	300	&	120\\
\ce{PIm+I-}		& 78	&	1.8	&	18	&	122	&	126	&	85	&	116	&	47\\
\ce{PPyr+Cl-}	& 270 	&	2.8	&	12	&	50.6	&	64	&	52	&	67	&	27\\
\ce{PPyr+Br-}	& 9800 	& 	0.3	&	2	&	1.15	&	2.8	&	5	&	7.3	&	2.9\\
\ce{PPyr+I-}	& 8800	&	0.0	&	89	&	1.08	&	2.5	&	4.6	&	7.2	&	2.9\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. $A$ and $E_c$ are obtained supposing $\tau=0.588$.}
\label{tab:results}
\end{table}

@@

For the theory to be correct, we need the electrostatic length to be larger than the thermal length, in other words $g_e \geq g_T$. Using Equations (\ref{eq:geD}) and (\ref{eq:B}), we express $g_e$ function of $B$ and $\tau$:
\begin{equation}
g_e = \left(\frac{B^3}{g_\mathrm{K}\tau}\right)^{1/2},
\end{equation}
which yields the condition $\tau \geq \frac{g_\mathrm{K}}{B}$. This condition it the most stringent for PPyrI (smallest $B$) and yields $\tau \geq 0.58$. In the following we set $\tau$ to this minimal value that corresponds to $\Theta=\SI{450}{\celsius}$.

The values we compute for $A$ using Equation (\ref{eq:A}) are shown in Table~\ref{tab:results}. $A$ varies between $4.6$ for \ce{PPyr+I-}, corresponding to $15$ uncondensed counterions per polymer, to $383$ for \ce{PIm+Br-} indicating less than a single dissociated counterion every 5 polymers.

\subsubsection{Extention of the electrostatic blobs}

@@

If we consider that the head-to-body bonds break at $\gamma_c$, we can extract their bonding energy $E_c$ from Equation (\ref{eq:stretch}). The results are reported in Table~\ref{tab:results}. In the most dissociated \ce{PPyr+I-} and \ce{PPyr+Br-}, the head-to-body bonding energy $E_c$ is compatible with typical ionic bonds, \SI{5}{\kilo\joule\per\mole} in water~\cite{Schneider1992}. However the bonding energy increases with increasing counterion condensation to reach 85\% of the carbon-carbon single bond (\SI{350}{\kilo\joule\per\mole} or $140 k_\mathrm{B}T$) in the case of \ce{PIm+Br-}. Indeed, at high counterion condensation, the electrostatic blobs are large, up to 530 polymers for \ce{PIm+Br-}, and embed head-to-body bonds into compact blobs of low water content. Such a low polarity microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly~\cite{Rehm2010}.

>>>>

\begin{figure}
\includegraphics{procession.pdf}
\caption{Processionary model. Sketch of the network in the case of a procession of size $n=3$ between cross-links (gray disks) and a persistence length $r_\mathrm{src}\approx 3D$. Empty circles are not individual monomers but electrostatic blobs of size $D$. Electrostatic blobs containing an anionic head are shown as black filled circles.}
\label{fig:dissociation}
\end{figure}

@@

\begin{table}
\sisetup{
separate-uncertainty,
%table-format = 1.3(1),
}
\begin{tabular}{@{}l|SSS|SSSSSS@{}}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_0$} & {$\gamma_c$} & {$n$} & {$B$} & {$E_c$} & {$E_c$} & {$D$} & {$A$}\\
&	{\si{\pascal}} & {\%} & {\%} & & & {\si{\kilo\joule\per\mole}}& {$k_\mathrm{B}T$} & {$\ell_B$}&\\\hline&&&&&\\[-10pt]
\ce{PIm+Br-}	& 13	&	1.1	&	810	&	860	&	930	&	210	&	84	&	13	&	460\\
\ce{PIm+I-}		& 78	&	1.8	&	18	&	122	&	126	&	80	&	32	&	5	&	103\\
\ce{PPyr+Cl-}	& 270 	&	2.8	&	12	&	50.6	&	64	&	46	&	19	&	3.4	&	62\\
\ce{PPyr+Br-}	& 9800 	& 	0.3	&	2	&	1.15	&	2.8	&	5.05	&	2.02	&	0.7	&	6\\
\ce{PPyr+I-}	& 8800	&	0.0	&	89	&	1.08	&	2.5	&	5	&	2	&	0.7	&	5.5\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. $E_c$, $D$ and $A$ are obtained supposing $\tau=0.46$.}
\label{tab:results}
\end{table}

@@

We know that $\Theta>\SI{100}{\celsius}$ so $0.2<\tau<1$. In the following we will deduce a more precise measure of $\tau$ and thus values of $A$.

%For the theory to be correct, we need the electrostatic length to be larger than the thermal length, in other words $g_e \geq g_T$. Using Equations (\ref{eq:geD}) and (\ref{eq:B}), we express $g_e$ function of $B$ and $\tau$:
%\begin{equation}
%g_e = \left(\frac{B^3}{g_\mathrm{K}\tau}\right)^{1/2},
%\end{equation}
%which yields the condition $\tau \geq \frac{g_\mathrm{K}}{B}$. This condition it the most stringent for PPyrI (smallest $B$) and yields $\tau \geq 0.58$. In the following we set $\tau$ to this minimal value that corresponds to $\Theta=\SI{450}{\celsius}$.
%
%The values we compute for $A$ using Equation (\ref{eq:A}) are shown in Table~\ref{tab:results}. $A$ varies between $4.6$ for \ce{PPyr+I-}, corresponding to $15$ uncondensed counterions per polymer, to $383$ for \ce{PIm+Br-} indicating less than a single dissociated counterion every 5 polymers.

\subsubsection{Extension of the electrostatic blobs}

@@

If we consider that the head-to-body bonds break at $\gamma_c$, we can express the reduced temperature as
\begin{equation}
\tau = B^{1/3} g_\mathrm{K}^{5/3} \left(\frac{1}{(\gamma_c +1)^{1/2} -1}\frac{E_c}{N k_\mathrm{B}T}\right)^{2/3}
\label{eq:tau}
\end{equation}
where $E_c$ is the energy of the head-to-body bond that consists in two ionic bonds. In water the bonding energy between two ions is typically \SI{5}{\kilo\joule\per\mole}\cite{Schneider1992}, thus $E_c\approx 2 k_\mathrm{B}T$. However the medium surrounding the ionic bond is not water. One has to keep in mind that the chain is in bad solvent, thus electrostatic blobs have low water content. Such a low polarity microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly\cite{Rehm2010}.

Using Equations~(\ref{eq:geD}) and (\ref{eq:B}) we express the size $D$ of electrostatic blobs function of $B$ and $\tau$:
\begin{equation}
g_e = \left(\frac{B^3}{g_\mathrm{K}\tau}\right)^{1/2},\qquad 
D = b\left(\frac{B}{g_\mathrm{K}^3\tau}\right)^{1/2}.
\label{eq:DB}
\end{equation}

We thus expect stronger head-to-body bonds when $B$ is large and the low-polarity environment is extended ($D\gg\ell_\mathrm{B}$). A contrario, the weakest ionic links, closest to their strength in water, should be found in \ce{PPyr+I-} where $B$ is minimum (hopefully $D<\ell_\mathrm{B}$).

Applying Equation~(\ref{eq:tau}) to \ce{PPyr+I-} with $E_c = 2 k_\mathrm{B}T$ yields $\tau \approx 0.46\pm 0.1$. Using this value and Equation (\ref{eq:stretch}) we deduce the bounding energy for each composition, as reported in Table~\ref{tab:results}. Indeed, the bonding energy increases with increasing counterion condensation to reach 60\% of the carbon-carbon single bond (\SI{350}{\kilo\joule\per\mole} or $140 k_\mathrm{B}T$) in the case of \ce{PIm+Br-}. According to Equation (\ref{eq:DB}), in this extreme case an electrostatic blob contains $\approx 500$ polymers which embed the head-to-body bonds in a low-polarity environment 13 times larger than the Bjerrum length. Conversely, for \ce{PPyr+I-} and \ce{PPyr+Br-} electrostatic blobs are smaller than $\ell_\mathrm{B}$, confirming an aqueous microenvironment. Quantitatively, the bonding energy of an ionic bond is inversely proportional to the relative dielectric constant $\epsilon_r$, therefore we need to suppose a local $\epsilon_r$ around 2 (typical of water-insoluble polymers) instead of 80 for water to recover the large bonding energy of \ce{PIm+Br-}.

Table~\ref{tab:results} also shows the values of $A$ deduced from Equation~(\ref{eq:A}) and the measured value of $\tau$. $A$ varies between $5.5$ for \ce{PPyr+I-}, corresponding to $13$ uncondensed counterions per polymer, to $460$ for \ce{PIm+Br-} indicating a single dissociated counterion every 7 polymers.

From purely rheological measurements, we have thus obtained microscopic information on the state of the gel: amount of charge condensation, head-to-body bounding energy and size of a procession.
